An Implementation of Reference Counting for a Python to x86 Compiler

### 2011-12-15 paper/paper.tex
tweak grammar and commas

There are currently two widely used strategies to perform automatic memory management.  One strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.  This set of objects includes all global objects, as well as the objects which are reachable from the current stack frame.  Any objects that are traversed during this operation are marked as reachable and therefore cannot be deallocated.  Any object which is not marked is a candidate for deallocation.  The other primary strategy for automated memory management is reference counting garbage collection. Reference counting works by keeping a count of references to a particular object.  References are tracked by incrementing and decrementing an object's reference count when an assignment to a pointer occurs or when a pointer variable goes out of scope.  When the reference count reaches zero, it is safe to deallocate the object.

@@

Because reference counting garbage collection can deallocate an object at the specific program point where there are no longer any references, it can allow for finalizers to be run immediately.  This makes the use of finalization a valid approach to releasing system resources.

@@

We start with an existing compiler based upon previous work \cite{siek}.  This compiler converts a subset of the Python language to x86 assembly and supports one of the core features of Python, namely that everything is a first class object.  Prior to our modifications, the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs with minimal memory constraints.  After our modifications, the compiler properly maintains reference counts for all objects created and automatically deallocates objects when there are no longer any references to them.

@@

\begin{itemize}
\item Objects must include a counter that tracks the number of references to that object.
\item Functions to increment and decrement an object's reference count must be created.
\item Assignment of an object to a pointer or \textit{reference} should result in correct updates to an object's reference count.
\item Reference counts must be decremented when a program reaches the end of a given scope, since local variables are destroyed.
\item When an object's reference count reaches zero, it is no longer reachable and can be safely destroyed.

@@

To support incrementing and decrementing an object's reference count, we create corresponding functions in our runtime called \afunction{inc\_ref\_ctr} and \afunction{dec\_ref\_ctr}.  The increment function is trivial in that it simply de-references the object pointer and increments the \afield{ref\_ctr} field.   However, the decrement function potentially has more work to do than just de-referencing the pointer and decrementing the reference count.  According to the steps for reference counting mentioned previously, we need to deallocate an object whenever its reference count reaches zero.  To provide this behavior, the \afunction{dec\_ref\_ctr} function has to check if the reference count has reached zero, and if so, delegate to a corresponding \afunction{free\_\textit{type}} method to free an object of the given type.

@@

In some cases, the type definition for an object in our runtime did not aggregate a \atype{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \atype{class\_struct}).  For this reason, significant changes were made to the runtime to ensure that any type that references another type has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.  To ensure that we did not make any mistakes in this process, we created a runtime test suite that used the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are deallocated when their reference count reaches zero.  This runtime test suite is important to unit test the runtime library before attempting to use it from our compiler.

Finally, in order to verify the correctness of our implementation, we create a set of abstractions on top of the system allocation functions \afunction{malloc} and \afunction{free}.  These abstractions are part of our runtime and provide us the following benefits:

@@

The abstractions are implemented in a new module appropriately called \texttt{pymem}.  Every allocation of an object in our runtime is modified to use the \afunction{pymem\_new} and \afunction{pymem\_free} instead of \afunction{malloc} and \afunction{free}. As a result, we can determine exactly how much memory has been allocated and what objects have still not been deallocated by calling the \afunction{pymem\_print\_statistics} function at any given program point.  In practice, this is only called at the end of each program\footnote{This is achieved by modifying the x86 generation phase of our compiler to include calls to set up our \texttt{pymem} module and print statistics at the end of the program. } to ensure that all memory has been deallocated.  Since we can track the time of allocation and deallocation, this module can be beneficial for investigating object lifetime as well.

@@

\subsubsection{Decrementing reference counts}
An assignment to a variable should result in a decrement of the reference count for the object referred to by the variable before the assignment.  However, this can't be done when the variable doesn't refer to anything, which is true prior to the initial assignment.  In fact, trying to even access such a variable before it is assigned produces undefined behavior since there is no guarantee on the contents of the variable (it could be garbage or a value from a previous stack frame).   To prevent a call to \afunction{dec\_ref\_ctr} on a variable before the initial assignment, we initially tried to statically determine when the first assignment occurred.  This approach failed when we realized that for some programs, it is not possible to determine where the initial assignment occurs at compile-time.  Listing~\ref{lst:initialassign} shows an example program where the initial assignment to the variable \avariable{x} cannot be determined statically.  If the input provided by the user evaluates to \aliteral{True}, then the initial assignment to $x$ occurs at line 2.  However, if the input evaluates to \aliteral{False}, then the initial assignment to \avariable{x} occurs at line 3.  If we transform the input program so that a call is made to decrement the reference count for \avariable{x} immediately prior to the assignment on line 3, then there is the possibility that this will cause the program to crash if the user provides an input that evaluates to \aliteral{False}.

To solve this issue, we make a simple change.  We ensure that the value of \avariable{x} is always initialized, by assigning it the value zero at the program scope where it is defined\footnote{Python defines this as the block where a variable is initially assigned to.}.  In our compiler, this is equivalent to a primitive of type \atype{int}, but this can be any sentinel value that results in a no-op when a call is made to \afunction{dec\_ref\_ctr}.  By ensuring that all local variables are assigned a sentinel value at the beginning of the scope where they are defined, we can safely transform our input program to call \afunction{dec\_ref\_ctr} immediately preceding an assignment, and \afunction{inc\_ref\_ctr} immediately succeeding an assignment.  Listing~\ref{lst:initialassign2} shows the results of this change.

This solution is somewhat controversial, since it can alter program behavior when the program tries to read from a variable that has not yet been assigned to.  The standard Python interpreter throws a \textit{NameError} in this case, but our compiler will allow such a program to be compiled.  Reading the variable will then result in the sentinel value (zero in our case) being read, instead of the program crashing or having undefined behavior.  A program which used uninitialized (or declared) variables would then have its runtime behavior changed.  Another option to address this issue, is to make it a compile-time error to try and access variable from a conditional block that may have not been previously initialized.  This is similar to a restriction in the Java language.

@@

Return statements are also tricky to get right.  Since return statements can appear in any order in the body of a function, we have to ensure that a return statement will not skip any decrement of reference counts associated with the end of the function scope.  To ensure that a return statement will execute this ``cleanup'' code, we add a function context that contains the list of local assignments and function arguments.  This context is available when processing any return nodes, and therefore the decrement of reference counts can be inserted prior to the return from the function.

Finally, return values must be accounted for.  Since an object may be created in a function and returned to the caller, we cannot decrement its reference count to zero before returning from the function, otherwise it will be immediately deallocated.  Our first heuristic to handle this removed the call to decrement the reference count for the variable being returned from a function.  Correspondingly, the caller did not increment its reference count after assigning a return value from a function.  However, this heuristic failed for the case where a function argument is returned.  In this case, the function argument has one additional reference count that is introduced at the beginning of the function, and therefore will never be decremented to zero.  Instead, we chose to implement a different approach by deferring the decrement of the reference count for any variable that is being returned from a function.  This is done by adding the \afunction{autorelease}\footnote{autorelease is borrowed from the Objective-C keyword with the same meaning} function to the runtime library.  This function will defer the decrement of a reference count for an object until the next call to \afunction{dec\_ref\_ctr}.  This is done by incrementing a new field, \afield{deferred\_cnt} in the C structure representing an object.  On the next call to \afunction{dec\_ref\_ctr}, this field is consulted to determine how many times to decrement the reference count.

\subsubsection{Other challenges}
In the runtime, a new list is created when two lists are added together.  Our compiler correctly handles reference counting for the left hand side and the right hand side of the addition (handled by the intermediate flatten phase), however the newly created list references the individual elements from both lists.  Therefore, the runtime had to be modified to correctly increment the reference counts for each of the elements from the original lists which are referenced in the new, concatenated list.

@@

Finally, heapification presents some unique challenges that require additional work to overcome.  Heapification occurs after explication, and therefore after our reference counting AST transformation.  This causes problems when the heapification phase introduces additional assignments to heapify variables.  These additional assignments are then never referenced counted correctly.  Also, the process of heapification encapsulates a variable in a list, and transforms assignments to that variable to assignments to a subscript.  This is problematic because our \afunction{set\_subscript} runtime function already handles incrementing and decrementing reference counts internally, so when an assignment that is surrounded by corresponding increment and decrement calls is converted to an assignment to a subscript, these increment and decrement calls are unnecessary.  We cannot move the reference counting AST transformation after the explication phase, as variable are not tagged with a type after this phase.  Therefore, to solve this issue, further work needs to done investigate if heapification can be done prior to explication.  Despite these issues, we found that heapification still works, but causes memory leaks.

@@

\subsection{Correctness}
We evaluated our reference counting implementation for correctness by using the \texttt{valgrind} utility against a simple application that uses a large number of language features.  This program is given in listing~\ref{lst:allfeatures}.  Without garbage collection, valgrind reports many leaks.  With reference counting, valgrind reports only a single leak for the same program.  The valgrind results are given in figure~\ref{fig:valgrind}.

@@

To measure the overhead associated with reference counting, we measure the average execution time of the sample program given in listing~\ref{lst:allfeatures} over ten runs.  The results are given in figure~\ref{fig:overhead}.  We see that our reference counting implementation imposes an overhead of nearly 650\% when compared to the compiler without memory management.

Obviously, there is room for improvement in our implementation.  One such improvement would be simple dispatch code to save a call to the runtime functions \afunction{inc\_ref\_ctr} and \afunction{dec\_ref\_ctr} when a variable contains a primitive.  Going one step further, a type specialization stage could infer the possible types of variables using static program analysis, and could help to further eliminate this dispatch code.  As mentioned in the related work, lazy deallocation can also help to batch deallocations and increase program throughput.

@@

The latency introduced by the \textit{dec\_ref\_ctr} is linear with the number of nodes in the object graph being deallocated.  To illustrate this, the program given in listing~\ref{lst:tree} is run with different inputs, where the input is the height of a 4-ary tree.  The number of nodes in such a tree is given by $\frac{{k}^{\left(h+1\right)}-1}{k-1}$ where $k=4$, and $h$ is the height of tree.  The results are graphed in figure~\ref{fig:treelatency}.  In typical programs, we expect the number of deallocations triggered by any given call to \textit{dec\_ref\_ctr} to be relatively small, but further evaluation of this for real world programs would be needed to prove this hypothesis.

@@

Reference counting is one of several options available to language designers who are looking for solutions to automated memory management.  We provide insight into the complexity of implementing such a feature, by extending an existing Python to x86 compiler to support reference counting garbage collection.  We find that while the premise behind reference counting is relatively simple, there are many nuances to integrating reference counting into such a compiler, especially in the presence of relatively complex language features such as closures and objects.

>>>>

There are currently two widely used strategies to perform automatic memory management.  One strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.  This set of objects includes all global objects as well as the objects which are reachable from the current stack frame.  Any objects that are traversed during this operation are marked as reachable and therefore cannot be deallocated.  Any object which is not marked is a candidate for deallocation.  The other primary strategy for automated memory management is reference counting garbage collection. Reference counting works by keeping a count of references to a particular object.  References are tracked by incrementing and decrementing an object's reference count when an assignment to a pointer occurs or when a pointer variable goes out of scope.  When the reference count reaches zero, it is safe to deallocate the object.

@@

Because reference counting garbage collection can deallocate an object at the specific program point where there are no longer any references it can allow for finalizers to be run immediately.  This makes the use of finalization a valid approach to releasing system resources.

@@

We start with an existing compiler based upon previous work \cite{siek}.  This compiler converts a subset of the Python language to x86 assembly and supports one of the core features of Python, namely that everything is a first class object.  Prior to our modifications the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs with minimal memory constraints.  After our modifications, the compiler properly maintains reference counts for all objects created and automatically deallocates objects when there are no longer any references to them.

@@

\begin{itemize}
\item Objects must include a counter that tracks the number of references to that object.
\item Functions to increment and decrement an object's reference count must be created.
\item Assignment of an object to a pointer or \textit{reference} should result in correct updates to an object's reference count.
\item Reference counts must be decremented when a program reaches the end of a given scope, since local variables are destroyed.
\item When an object's reference count reaches zero it is no longer reachable and can be safely destroyed.

@@

To support incrementing and decrementing an object's reference count, we create corresponding functions in our runtime called \afunction{inc\_ref\_ctr} and \afunction{dec\_ref\_ctr}.  The increment function is trivial in that it simply de-references the object pointer and increments the \afield{ref\_ctr} field.   However, the decrement function potentially has more work to do than just de-referencing the pointer and decrementing the reference count.  According to the steps for reference counting mentioned previously, we need to deallocate an object whenever its reference count reaches zero.  To provide this behavior, the \afunction{dec\_ref\_ctr} function has to check if the reference count has reached zero and, if so, delegate to a corresponding \afunction{free\_\textit{type}} method to free an object of the given type.

@@

In some cases the type definition for an object in our runtime did not aggregate a \atype{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \atype{class\_struct}).  For this reason significant changes were made to the runtime to ensure that any type that references another type has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.  To ensure that we did not make any mistakes in this process we created a runtime test suite that used the abstractions provided by the runtime library to create objects, increment and decrement their reference counts, and ensure that all objects are deallocated when their reference count reaches zero.  This runtime test suite is important to unit test the runtime library before attempting to use it from our compiler.

Finally, in order to verify the correctness of our implementation we create a set of abstractions on top of the system allocation functions \afunction{malloc} and \afunction{free}.  These abstractions are part of our runtime and provide us the following benefits:

@@

The abstractions are implemented in a new module appropriately called \texttt{pymem}.  Every allocation of an object in our runtime is modified to use the \afunction{pymem\_new} and \afunction{pymem\_free} instead of \afunction{malloc} and \afunction{free}. As a result, we can determine exactly how much memory has been allocated and what objects have still not been deallocated by calling the \afunction{pymem\_print\_statistics} function at any given program point.  In practice this is only called at the end of each program\footnote{This is achieved by modifying the x86 generation phase of our compiler to include calls to set up our \texttt{pymem} module and print statistics at the end of the program. } to ensure that all memory has been deallocated.  Since we can track the time of allocation and deallocation, this module can be beneficial for investigating object lifetime as well.

@@

\subsubsection{Decrementing reference counts}
An assignment to a variable should result in a decrement of the reference count for the object referred to by the variable before the assignment.  This cannot, however, be done when the variable does not refer to an allocated memory region, which is true prior to the initial assignment.  In fact, trying to even access such a variable before it is assigned produces undefined behavior since there is no guarantee on the contents of the variable (it could be garbage or a value from a previous stack frame).   To prevent a call to \afunction{dec\_ref\_ctr} on a variable before the initial assignment, we initially tried to statically determine when the first assignment occurred.  This approach failed when we realized that for some programs it is not possible to determine where the initial assignment occurs at compile-time.  Listing~\ref{lst:initialassign} shows an example program where the initial assignment to the variable \avariable{x} cannot be determined statically.  If the input provided by the user evaluates to \aliteral{True}, then the initial assignment to $x$ occurs at line 2.  However, if the input evaluates to \aliteral{False}, then the initial assignment to \avariable{x} occurs at line 3.  If we transform the input program so that a call is made to decrement the reference count for \avariable{x} immediately prior to the assignment on line 3, then there is the possibility that this will cause the program to crash if the user provides an input that evaluates to \aliteral{False}.

To solve this issue it was necessary to make a simple change: ensure that the value of \avariable{x} is always initialized by assigning it the value zero at the program scope where it is defined\footnote{Python defines this as the block where a variable is initially assigned to.}.  In our compiler, this is equivalent to a primitive of type \atype{int}, but this can be any sentinel value that results in a no-op when a call is made to \afunction{dec\_ref\_ctr}.  By ensuring that all local variables are assigned a sentinel value at the beginning of the scope where they are defined, we can safely transform our input program to call \afunction{dec\_ref\_ctr} immediately preceding an assignment, and \afunction{inc\_ref\_ctr} immediately succeeding an assignment.  Listing~\ref{lst:initialassign2} shows the results of this change.

This solution is not ideal since it can alter program behavior when the program tries to read from a variable that has not yet been assigned a value.  The standard Python interpreter throws a \textit{NameError} in this case, but our compiler will allow such a program to be compiled.  Reading the variable will then result in the sentinel value (zero in our case) being read, instead of the program crashing or having undefined behavior.  A program which used uninitialized (or declared) variables would then have its runtime behavior changed.  Another option to address this issue, is to make it a compile-time error to try and access variable from a conditional block that may have not been previously initialized.  This is similar to a restriction in the Java language.

@@

Second, \afunction{return} statements are also tricky to get right.  Since return statements can appear in any order in the body of a function, we have to ensure that a \afunction{return} statement will not skip any decrement of reference counts associated with the end of the function scope.  To ensure that a \afunction{return} statement will execute this ``cleanup'' code, we add a function context that contains the list of local assignments and function arguments.  This context is available when processing any return nodes, and therefore the decrement of reference counts can be inserted prior to the return from the function.

Finally, return values must be accounted for.  Since an object may be created in a function and returned to the caller we cannot decrement its reference count to zero before returning from the function, otherwise it will be immediately deallocated.  Our first heuristic to handle this removed the call to decrement the reference count for the variable being returned from a function.  Correspondingly, the caller did not increment its reference count after assigning a return value from a function.  However, this heuristic failed for the case where a function argument is returned.  In this case, the function argument has one additional reference count that is introduced at the beginning of the function, and therefore will never be decremented to zero.  Instead, we chose to implement a different approach by deferring the decrement of the reference count for any variable that is being returned from a function.  This is done by adding the \afunction{autorelease}\footnote{autorelease is borrowed from the Objective-C keyword with the same meaning} function to the runtime library.  This function will defer the decrement of a reference count for an object until the next call to \afunction{dec\_ref\_ctr}.  This is done by incrementing a new field, \afield{deferred\_cnt} in the C structure representing an object.  On the next call to \afunction{dec\_ref\_ctr} this field is consulted to determine how many times to decrement the reference count.

\subsubsection{Other challenges}
In the runtime a new list is created when two lists are added together.  Our compiler correctly handles reference counting for the left hand side and the right hand side of the addition (handled by the intermediate flatten phase), however the newly created list references the individual elements from both lists.  Therefore, the runtime had to be modified to correctly increment the reference counts for each of the elements from the original lists which are referenced in the new, concatenated list.

@@

Finally, heapification presents some unique challenges that require additional work to overcome.  Heapification occurs after explication, and therefore after our reference counting AST transformation.  This causes problems when the heapification phase introduces additional assignments to heapify variables.  These additional assignments are then never referenced counted correctly.  Also, the process of heapification encapsulates a variable in a list, and transforms assignments to that variable to assignments to a subscript.  This is problematic because our \afunction{set\_subscript} runtime function already handles incrementing and decrementing reference counts internally; so when an assignment that is surrounded by corresponding increment and decrement calls is converted to an assignment to a subscript these increment and decrement calls are redundant.  We cannot move the reference counting AST transformation after the explication phase, as variable are not tagged with a type after this phase.  Therefore, to solve this issue, further work needs to done investigate if heapification can be done prior to explication.  Despite these issues we found that heapification still works, but may cause memory leaks.

@@

\subsection{Correctness}
We evaluated our reference counting implementation for correctness by using the \texttt{valgrind} utility against a simple application that uses a large number of language features.  This program is given in listing~\ref{lst:allfeatures}.  Without garbage collection \texttt{valgrind} reports many memory leaks.  With reference counting \texttt{valgrind} reports only a single leak for the same program.  The \texttt{valgrind} results are given in figure~\ref{fig:valgrind}.

@@

To measure the overhead associated with reference counting we measure the average execution time of the sample program given in listing~\ref{lst:allfeatures} over ten runs.  The results are given in figure~\ref{fig:overhead}.  We see that our reference counting implementation imposes an overhead of nearly 650\% when compared to the compiler without memory management.

There is obvious room for improvement in our implementation.  One such improvement would be simple dispatch code to save a call to the runtime functions \afunction{inc\_ref\_ctr} and \afunction{dec\_ref\_ctr} when a variable contains a primitive.  Going one step further, a type specialization stage could infer the possible types of variables using static program analysis, and could help to further eliminate this dispatch code.  As mentioned in the related work, lazy deallocation can also help to batch deallocations and increase program throughput.

@@

The latency introduced by the \textit{dec\_ref\_ctr} is linear with the number of nodes in the object graph being deallocated.  To illustrate this the program given in listing~\ref{lst:tree} is run with different inputs, where the input is the height of a 4-ary tree.  The number of nodes in such a tree is given by $\frac{{k}^{\left(h+1\right)}-1}{k-1}$ where $k=4$, and $h$ is the height of tree.  The results are graphed in figure~\ref{fig:treelatency}.  In typical programs, we expect the number of deallocations triggered by any given call to \textit{dec\_ref\_ctr} to be relatively small, but further evaluation of this for real world programs would be needed to prove this hypothesis.

@@

Reference counting is one of several options available to language designers who are looking for solutions to automated memory management.  We provide insight into the complexity of implementing such a feature by extending an existing Python to x86 compiler to support reference counting garbage collection.  We find that while the premise behind reference counting is relatively simple, there are many nuances to integrating reference counting into such a compiler, especially in the presence of relatively complex language features such as closures and objects.
